---
title: "Species Frequency & Detection Probability Estimates"
author: "Matthew Morgan"
date: "2024-10-16"
output:
  html_document: default
  pdf_document: default
---

This code prepares data for further analysis and merges the master bird DB 
with transect visibility and bird details (including mass, trophic niche and 
trophic position). VIF is tested with LM to test for collinearity. Multiple 
models are compared with various combinations and interactions of covariates. 
Code is setup to adjust master db with detection functions.  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(Distance)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
library(readr)
```

```{r set directory}
setwd("C:/Users/721049/OneDrive - hull.ac.uk/Bird_DA_GitHub")
getwd()
```

```{r import data}
bird <- read_csv("Master_DB_Birds.csv", 
                show_col_types = FALSE)
sum(bird$count) # 14708

detect <- read_csv("detectability_category.csv", 
                 show_col_types = FALSE)

species_size <- read_csv("tran.birdlist.alldetails.csv")
```
## Composition of bird speicies

```{r bird encounter frequency}
# Summarise data by species, calculate total count, add frequency, and percentage column
bird_summary <- bird %>%
  group_by(species) %>%                                   
  summarise(
    total_count = sum(count),                            
    frequency_detected = n()                              
  ) %>%
  mutate(percentage = (total_count / sum(total_count)) * 100) 

#
print(bird_summary)
write_csv(bird_summary, "bird_count_freq.csv")
```

Calculate bird data Yes and No Freq with %
```{r bird encounter with and without water}
# Summarise data by species, calculate total count, include water presence details, add frequency, and percentage column
bird_summary <- bird %>%
  group_by(species) %>%
  summarise(
    total_count = sum(count),                               
    water_present_yes = sum(water == "Yes", na.rm = TRUE),  
    water_present_no = sum(water == "No", na.rm = TRUE),    
    frequency_detected = n()                               
  ) %>%
  mutate(
    percentage = (total_count / sum(total_count)) * 100,          
    water_yes_percentage = (water_present_yes / frequency_detected) * 100,  
    water_no_percentage = (water_present_no / frequency_detected) * 100     
  )

# 
print(bird_summary)

# save
write_csv(bird_summary, "bird_count_freq_water_presence.csv")


```

```{r bird count summary}
# Summarise data by species
bird_count <- bird %>%
  group_by(species) %>%
  summarise(
    total_count = sum(count),                               
    water_present_yes = sum(count[water == "Yes"], na.rm = TRUE), 
    water_present_no = sum(count[water == "No"], na.rm = TRUE)    
  )

# 
print(bird_count)
sum(bird_count$total_count)
# 
write_csv(bird_count, "bird_count_water_presence.csv")
```

## Detection probability estimates

```{r merge data}
# Merging bird and detect based on the "transect and transect_section
merged_data <- merge(bird, detect, by = c("transect", "transect_section"))
colnames(merged_data)[23] <- "Region.Label"

#
head(merged_data)
unique(merged_data$species) # 79
sum(merged_data$count) # 14708
```

```{r correct names}
colnames(merged_data)[colnames(merged_data) == "species"] <- "common_name"

merged_data <- merge(merged_data, species_size, by = c("common_name"))

# View the merged data
head(merged_data)
 unique(merged_data$common_name) # 79
 sum(merged_data$count) # 14708
```

Filter birds from bands 3-4 they have unbounded distances which makes it impossible
to use for detectability estimates. Mid-point distances added (0-25 =12.5, 25-100 = 62.5)
```{r filter to distance bands 1 and 2}
filtered_data <- merged_data[merged_data$distance_code %in% c(1, 2), ]

# Create a numeric 'distance' column based on the 'distance_code' (mid-point)
filtered_data$Distance <- ifelse(filtered_data$distance_code == 1, 12.5, 62.5)

# Rename the "observer" column to avoid conflict
# colnames(filtered_data)[colnames(filtered_data) == "observer"] <- "observer_name"

# Check if the column was renamed successfully
head(filtered_data)
```

```{r check data availability for modelling}
table(filtered_data$Region.Label)

# Summarise counts by species, distance, and habitat
species_distance_summary <- filtered_data %>%
  group_by(size_class, Distance, Region.Label) %>%
  summarise(count = n(), .groups = 'drop')

#
sum(species_distance_summary$count)
```

Plot species detection declines
```{r species detection declines}
# Create column to represent distance with values 1 and 2
species_distance_summary$Distance.Bands <- ifelse(species_distance_summary$Distance == 12.5, 1, 2)

# Set the order of Region.Label to be 'open', 'semi', 'closed'
species_distance_summary$Region.Label <- factor(species_distance_summary$Region.Label, levels = c("open", "semi", "closed"))

# Colours for each size class
manual_colors <- c("large" = "#1f77b4", "medium" = "#ff7f0e", "small" = "#2ca02c")  

#
ggplot(species_distance_summary, aes(x = Distance.Bands, y = count, color = size_class, group = size_class)) +
  geom_point(size = 4) +         
  geom_line(size = 1) +          
  labs(
    title = "Count vs Distance Across Habitats",
    x = "Distance Bands",        
    y = "Count",
    color = "Size Class"
  ) +
  scale_x_continuous(breaks = c(1, 2), labels = c("1", "2")) +  # Change x-axis to 1 and 2
  scale_color_manual(values = manual_colors) +   
  facet_wrap(~ Region.Label) +                   
  theme_minimal() +                              
  theme(
    plot.title = element_text(hjust = 0.5, size = 14), 
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 13)
  )

```

Create summary table to look at the amount of data available for modelling
```{r data available for modelling}
# Summary table
summary_table <- species_distance_summary %>%
  group_by(size_class, Region.Label) %>%
  summarise(
    Count_Band_1 = sum(count[Distance.Bands == 1]),   
    Count_Band_2 = sum(count[Distance.Bands == 2]),   
    Total_Count = sum(count)                         
  ) %>%
  ungroup()

# View the summary
summary_table
sum(summary_table$Total_Count) #4065
```

```{r histograms of birds sizes and habitats divided by water/no water}
str(species_distance_summary)
# Group data by transect and transect section
summary_1 <- filtered_data %>%
  group_by(transect, transect_section, size_class, water, Region.Label) %>%
  summarize(count = n())

# Print the summarized data
print(summary_1)

#
ggplot(summary_1, aes(x = Region.Label, fill = water)) +
  geom_bar(position = "dodge") +
  labs(title = "Habitat Distribution across transects",
       x = "Size Class",
       y = "Count") +
  theme_minimal()

```

Running all models with all covariates to check for collinearity. 

```{r VIF check for collinearity}
table(filtered_data$size_class, filtered_data$Region.Label)

# Fit linear models to assess VIF with the added 'water' variable
model1 <- lm(Distance ~ size_class + Region.Label + observer + water, data = filtered_data)
model2 <- lm(Distance ~ size_class * Region.Label + observer + water, data = filtered_data)
model3 <- lm(Distance ~ size_class * Region.Label * observer + water, data = filtered_data)

# Calculate VIF for all predictors and interactions in the model
vif(model1, type = "predictor")
vif(model2, type = "predictor")
vif(model3, type = "predictor")

```

```{r data preparation}
str(filtered_data)
summary(filtered_data)

# correct df
convunit <- convert_units("meter", "kilometer", "hectare")
names(filtered_data)[names(filtered_data) == "Distance"] <- "distance"
filtered_data$distbegin <- ifelse(filtered_data$distance == 12.5, 0, 25)
filtered_data$distend <- ifelse(filtered_data$distance == 12.5, 25, 100)

# Remove the distance column from the data
filtered_data$distance <- NULL
filtered_data$distance_code <- NULL
names(filtered_data)[names(filtered_data) == "observer"] <- "OBs"
filtered_data
sum(filtered_data$count)
```
## Individual models

```{r Filtered no covariates}
ahn.0 <- ds(filtered_data, transect="line", key="hn",
truncation=100, convert_units = conv)
print(ahn.0$ddf$criterion)
print(ahn.0$ddf$meta.data$breaks)
```

```{r Filtered + Water}
ahn.w <- ds(filtered_data, transect="line", key="hn", formula=~water,
truncation=100, convert_units = conv)
print(ahn.w$ddf$criterion)
print(ahn.w$ddf$meta.data$breaks)
```

```{r Filtered + Size}
ahn.size <- ds(filtered_data, transect="line", key="hn", formula=~size_class,
truncation=100, convert_units = conv)
print(ahn.size$ddf$criterion)
print(ahn.size$ddf$meta.data$breaks)
summary(ahn.size)
```

```{r Filtered + habitat} 
ahn.hab <- ds(filtered_data, transect="line", key="hn", formula=~Region.Label,
truncation=100, convert_units = conv)
print(ahn.hab$ddf$criterion)
print(ahn.hab$ddf$meta.data$breaks)
```

```{r Filtered + Observer}
# Fit the first model (half-normal)
ahn.obs <- ds(filtered_data, transect="line", key="hn", formula=~OBs, truncation=100, convert_units = conv)
print(ahn.obs$ddf$criterion) 
print(ahn.obs$ddf$meta.data$breaks)
```

```{r Filtered + Size + Habitat}
ahn.size.hab <- ds(filtered_data, transect="line", key="hn", formula=~size_class+Region.Label,
truncation=100, convert_units = conv)
print(ahn.size.hab$ddf$criterion)
print(ahn.size.hab$ddf$meta.data$breaks)
```

```{r Filtered + Size + Habitat + Water}
ahn.size.hab.w <- ds(filtered_data, transect="line", key="hn", formula=~size_class+Region.Label+water,
truncation=100, convert_units = conv)
print(ahn.size.hab.w$ddf$criterion)
print(ahn.size.hab.w$ddf$meta.data$breaks)
```

```{r Filtered + Size + Observer}
ahn.size.obs <- ds(filtered_data, transect="line", key="hn", formula=~size_class+OBs,
truncation=100, convert_units = conv)
print(ahn.size.obs$ddf$criterion)
print(ahn.size.obs$ddf$meta.data$breaks)
```

```{r Filtered + Size + Observer + Water}
ahn.size.obs.w <- ds(filtered_data, transect="line", key="hn", formula=~size_class+OBs+water,
truncation=100, convert_units = conv)
print(ahn.size.obs.w$ddf$criterion)
print(ahn.size.obs.w$ddf$meta.data$breaks)
```


```{r Filtered + Habitat + Observer}
ahn.hab.obs <- ds(filtered_data, transect="line", key="hn", formula=~Region.Label+OBs,
truncation=100, convert_units = conv)
print(ahn.hab.obs$ddf$criterion)
print(ahn.hab.obs$ddf$meta.data$breaks)
```

```{r Filtered + Habitat + Observer + Water}
ahn.hab.obs.w <- ds(filtered_data, transect="line", key="hn", formula=~Region.Label+OBs+water,
truncation=100, convert_units = conv)
print(ahn.hab.obs$ddf$criterion)
print(ahn.hab.obs$ddf$meta.data$breaks)
```


```{r Filtered + Habitat + Observer + Size}
ahn.hab.obs.size <- ds(filtered_data, transect="line", key="hn", formula=~Region.Label+OBs+size_class,
truncation=100, convert_units = conv)
print(ahn.hab.obs.size$ddf$criterion)
print(ahn.hab.obs.size$ddf$meta.data$breaks)
```

```{r Filtered + Habitat + Observer + Size + Water}
ahn.hab.obs.size.w <- ds(filtered_data, transect="line", key="hn", formula=~Region.Label+OBs+size_class+water,
truncation=100, convert_units = conv)
print(ahn.hab.obs.size.w$ddf$criterion)
print(ahn.hab.obs.size.w$ddf$meta.data$breaks)
```


```{r Filtered + Size * Habitat}
ahn.size.xhab <- ds(filtered_data, transect="line", key="hn", formula=~size_class*Region.Label,
truncation=100, convert_units = conv)
print(ahn.size.xhab$ddf$criterion)
print(ahn.size.xhab$ddf$meta.data$breaks)
```

```{r Filtered + Size * Habitat * Water}
ahn.size.xhab.w <- ds(filtered_data, transect="line", key="hn", formula=~size_class*Region.Label*water,
truncation=100, convert_units = conv)
print(ahn.size.xhab.w$ddf$criterion)
print(ahn.size.xhab.w$ddf$meta.data$breaks)
```

```{r Summaries of Water Transect Models}
model_summary <- summarize_ds_models(ahn.0, ahn.w, ahn.size, ahn.hab, ahn.obs, ahn.size.hab, ahn.size.hab.w, ahn.size.obs, ahn.size.obs.w, ahn.hab.obs, ahn.hab.obs.w, ahn.hab.obs.size, ahn.hab.obs.size.w, ahn.size.xhab, ahn.size.xhab.w)

summary(ahn.size.xhab)

gof_ds(ahn.size.xhab)

# write to CSV in your working directory
write.csv(model_summary, "model_summary.csv", row.names = FALSE)

```

```{r Plots of best water models}
plot(ahn.size.xhab, nc=20, main="Size * Habitat", pch=20, pdf=TRUE)
plot(ahn.size.hab)
```

```{r best water model}
# Fit best model
ahn.best <- ds(filtered_data, transect="line", key="hn", formula=~size_class*Region.Label,
truncation=100, convert_units = conv)

# Plot model with OBS and MAS
plot(ahn.best, nc=10, main="Model with Size and Habitat Interaction",
pch=".", pdf=TRUE)
#
gof_result <- gof_ds(ahn.best)
# Print the results
print(gof_result)

```

Extracting detection probabilites by distances from best model. Note that the 
detection probability has been averaged and therefore values at 12.5 are the 
same as 62.5.
```{r extract probability estimates for habitat}
# Create a data frame for each combination of size_class, Region.Label, and distance
new_data <- expand.grid(
  size_class = c("small", "medium", "large"),
  Region.Label = c("closed", "semi", "open"),
  distance = c(12.5, 62.5)  # Midpoints of your distance bands
)

# Generate predictions using the model
predicted_probs <- predict(ahn.size.xhab, newdata = new_data)

# Extract the first value from list and convert to numeric values
predicted_probs <- sapply(predicted_probs, function(x) if (is.list(x)) x[[1]] else x)

# Add the predictions to new_data for easy reference
new_data$Detection_Probability <- predicted_probs

# Print the prediction results
print(new_data)

# Optional: Plot detection probabilities by size class and Region.Label
p <- ggplot(new_data, aes(x = distance, y = Detection_Probability, color = size_class)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Region.Label) +
  labs(title = "Detection Probability by Size Class and Transect Visibility",
       x = "Distance (m)", y = "Detection Probability") +
  theme_classic()

p

ggsave("detection_probability_plot.png", plot = p, width = 11.7, height = 8.3, units = "in", dpi = 300)
```
Extracting detection probabilities for small, medium, large birds across 
different habitats
```{r extract probability estimates for habitat and size}
# Define a function to create detection function plots for each region
plot_detection_function <- function(region_label, main_title, show_legend = TRUE, show_xlab = TRUE, show_ylab = TRUE, main_text_size = 1.5, axis_text_size = 1.2, legend_text_size = 1.2) {
  # Plot 
  plot(
    0, 0, type = "n",              # Create an empty plot
    xlim = c(0, 100),              # Distance range
    ylim = c(0, 1),                # Detection probability range up to 1
    main = main_title,             # Main title
    xlab = if (show_xlab) "Distance" else "",  # Conditional x-axis label
    ylab = if (show_ylab) "Detection Probability" else "",  # Conditional y-axis label
    cex.main = main_text_size,     # Main title size
    cex.axis = axis_text_size,     # Axis text size
    cex.lab = axis_text_size       # Axis label size
  )
  
  # Add detection probability curves for each size class
  add.df.covar.line(ahn.size.xhab, data = data.frame(size_class = "small", Region.Label = region_label), 
                    lwd = 3, col = "blue")
  add.df.covar.line(ahn.size.xhab, data = data.frame(size_class = "medium", Region.Label = region_label), 
                    lwd = 3, col = "green")
  add.df.covar.line(ahn.size.xhab, data = data.frame(size_class = "large", Region.Label = region_label), 
                    lwd = 3, col = "red")
  
  # Add legend only if show_legend is TRUE
  if (show_legend) {
    legend("topright", legend = c("Small", "Medium", "Large"),
           col = c("blue", "green", "red"), lwd = 3, cex = legend_text_size)
  }
}

# Open a PNG device to save the multi-plot layout
png("Filtered_Detection_Function_Comparison_simple.png", width = 3508, height = 1300, res = 300)

# Set up 1-row 3-column layout for the plots
par(mfrow = c(1, 3))

# Create the detection function plots for each region
plot_detection_function("closed", "Closed Transect", show_legend = FALSE, show_xlab = FALSE, show_ylab = TRUE)
plot_detection_function("semi", "Semi-closed Transect", show_legend = FALSE, show_xlab = TRUE, show_ylab = FALSE)
plot_detection_function("open", "Open Transect", show_legend = TRUE, show_xlab = FALSE, show_ylab = FALSE)

# Reset layout after plotting
par(mfrow = c(1, 1))

# Close the PNG device to save the file
dev.off()


```


```{r}
knitr::kable(summarize_ds_models(ahn.0, ahn.w, ahn.size, ahn.hab, ahn.obs, ahn.size.hab, ahn.size.hab.w, ahn.size.obs, ahn.size.obs.w, ahn.hab.obs, ahn.hab.obs.w, ahn.hab.obs.size, ahn.hab.obs.size.w, ahn.size.xhab, ahn.size.xhab.w), 
             caption="Candidate model set for covariate analysis.",
             row.names = FALSE, digits=5)
```


adjusting the bird data base
```{r}
adjusted <- filtered_data

# Define correction factors
correction_factors <- data.frame(
  Region = c("closed", "closed", "closed", "semi", "semi", "semi", "open", "open", "open"),
  Size = c("small", "medium", "large", "small", "medium", "large", "small", "medium", "large"),
  Factor = c(0.163, 0.258, 0.289, 0.179, 0.238, 0.295, 0.225, 0.269, 0.297)
)

# use correction factors
adjusted <- adjusted %>%
  left_join(correction_factors, by = c("Region.Label" = "Region", "size_class" = "Size")) %>%
  mutate(
    true_counts = count / Factor
  )

print(adjusted)

#

write_csv(adjusted, "corrected_density.csv")


```


